---
  title: A4.  Factor and Chronological Classes, Advanced Ordering
---

```{r Environment set-up, include= FALSE}
remove( list= objects() )
```

```{r Document-Wide Knitr Options, include= FALSE }
knitr::opts_chunk$set(fig.width=6, fig.height=6) 
knitr::opts_chunk$set(comment= "")
knitr::opts_chunk$set(prompt= TRUE)
```

[Go to the Table of Contents](00_TableOfContents.html)

[Return to Last Lesson (A3: Classes, Constants, and Comparison Operators)](03_LessonA3.html)

[Proceed to Next Lesson (A5: Regular Expression)](05_LessonA5.html)

This lesson introduces categorical ("factor") and chronological ("POSIXct") classes of data.  It also covers more advanced material for object inspection, vector indexing, and statistical distributions.

### Object Inspection

You will need to inspect objects to ensure that a script is performing as intended.  There are at least three basic strategies for inspecting objects.  First, you can express the object in its entirety.  While this is effective for small objects, it becomes impractical for calculations at scale.  Second, you can express some part of the object.  This is effective for quick inspection to determine that an object has the expected form, changed from its earlier contents in expected ways, and has not experienced any unexpected coercion or NA propagation.  Here are some examples of how to inspect objects via partial expression.

```{R Partial Expression as Inspection}
## EXPRESSING PART OF THE OBJECT ==========

## The functions head() and tail() will express the top and bottom N cases
  ## respectively.  N defaults to 6, but you can use the n= argument to change N

head(letters, n= 8)

tail(letters, n= 8)


## Random sampling can be a powerful approach for expressing parts of
	## objects.  It gives you a better chance of spotting errors 
	## that only affect some elements of your vector, especially if
	## you do the inspection several times without using set.seed()

letters[
	sample(
		x= 1:length(letters),
		size= 8
		)
	]
```

Third, you can express some summary statistics about the object.  This is effective for gaining an abstract perspective on large scale objects or for checking for specific risks.  The class-sensitive function `summary()` is a great first step.  It displays various kinds of relevant summary information for different classes of objects.  In addition to `summary()` you can (and should) make testing functions as needed.  It falls to the programmer to determine what array of tests is enough to confirm that code works as intended.  To write error-free code, a programmer needs to be suspicious and skeptical.  Testing is a crucial part of this mindset.

```{R Summary Functions as Inspection}
## CREATE DEMONSTRATION OBJECTS ==========

numeric_object <- {1:10}^2
character_object <- letters[1:10]
logical_object <- {numeric_object %/% 2} == 0


## USING summary() TO DO OBJECT INSPECTION ==========

## The workhorse summary function in R is called summary().  Summary
	## is heavily class-sensitive, so it will produce different kinds
	## of summary outputs for different classes of objects.

## For numeric vectors, summary() will generate statistical moments.  You can
  ## also generate this using quantile() and mean().

summary(numeric_object)

mean(numeric_object)
quantile(numeric_object, probs= c( 0, 0.25, 0.5, 0.75, 1 ))

## For character vectors, summary() will fall back to generic information about
  ## the object.  You can also generate this information using length(),
  ## class(), and mode()
 
summary(character_object)

length(character_object)
mode(  character_object)
class( character_object)

## For logical vectors, summary() will generate similar information
	## but break down the length() counts by each possible value.

summary(logical_object)

## DEVISING OTHER TESTS TO DO OBJECT INSPECTION ==========

## This expressions calculates the percentage of invalid/missing
	## elements, i.e., the percentage of NA.  If this value exceeds zero
	## when there is no reason for NA elements to exist, or when this
	## value rises rapidly after a particular line of code, it is
	## a sign that you should investigate further.

mean( is.na(numeric_object) )

## This expression tests for a known distributional property of the
	## data.  Here we compare the actual frequency of different
	## groups of letters to the probability argument used in sample()
	## to generate the object, finding them roughly identical.

letter_frequency <- c(
	sum( character_object %in% letters[10:17] ),
	sum( character_object %in% letters[c( 1:4, 23:26 )] )
	)

input_frequency <- abs(1:26 - 13.5)
input_frequency <- input_frequency / sum(input_frequency)
input_frequency <- c(
	sum( input_frequency[10:17] ),
	sum( input_frequency[c(1:4, 23:26)] )
	)

round(letter_frequency / length(character_object) , digits= 3 )
round(input_frequency, digits= 3)

## This pair of expressions measures the percent all letters of
	## the alphabet are represented and the percent of all vector
	## elements that are letters.  Both of these should be one by
	## definition.  Anything else would indicate that something
	## unexpected has occurred.

mean(character_object %in% letters)
mean(letters %in% character_object)

## Delete demonstration objects that are no longer needed
remove(numeric_object, character_object, logical_object,
	input_frequency, letter_frequency)
```

### Factor and POSIXct Class - Working with Categorical and Chronological Data

In addition to the big three – numeric, character and logical – R has many other classes of vectors.  This lesson will address the class for categorical data ("factors") and a class for chronological data (POSIXct).  Both classes are extensions of big three vector modes.  Factors imbue characters with added meaning and chronological classes align quantities of seconds to the highly irregular increments (days, months, years, etc.) in which the passage of time is measured.

#### Factor Class - Managing Categorical Data

```{R Factor Class}
## Generate a simple character vector containing size categories
set.seed(1856)
character_vector <- sample(
	c("Small", "Medium", "Large"),
	size= 2^10,
	replace= TRUE
	)

head(character_vector)
summary(character_vector)

## Convert to factor class - R now understands that these are
	## categories and R summarizes them as such.
factor_vector <- as.factor(character_vector)

head(factor_vector)
summary(factor_vector)

## R can also generate factors by dividing numeric data into bins

shirt_neck <- seq(from= 14, to= 17.5, by= 0.5)
shirt_neck

shirt_neck <- cut(
	shirt_neck,
	breaks= c(-Inf, 15, 16, 17, Inf),
	right= FALSE,
	labels= c("S", "M", "L", "XL")
	)
shirt_neck

## Note how the object now expresses with the line "Levels: 
	## Large Medium Small" below the vector.  This is where R 
	## expresses the valid set of categories ("levels") for that
	## vector.  The vector will not accept categoricals that have
	## not first been added as a valid level. The levels() function
	## provides access to the valid category list.

shirt_neck[2] <- "QQ"
shirt_neck

levels(shirt_neck) <- c("S", "M", "L", "XL", "QQ")
shirt_neck[2] <- "QQ"
shirt_neck

## nlevels() provides a count of the number of valid categories,
	## confirming that a new category has been added.
nlevels(shirt_neck)

## Delete demonstration objects that are no longer needed
remove(shirt_neck, factor_vector, character_vector)
```

Factor class can be difficult to master because, compared to the big three, what you see is not what you have.  Understanding the quirks of factor class behavior is key to successful use.

```{R Factor Class Behavior}
## generate factor object
factor_vector <- factor(
	sample(
		x= c( "S", "L", "XL" ),
		size= 2^10, replace= TRUE
		)
	)
summary(factor_vector)

## One of the key features of factor class is that the object will
	## heed information about the factors contained. For instance, it will
	## recognize categories of data that are not present and also 
	## respect the ordering of categories.  In the example below,
	## medium is added as a valid category, and the categories are
	## reordered from smallest to largest.

factor_vector <- factor(
	factor_vector,
	levels= c("S", "M", "L", "XL")
	)
summary( factor_vector )

## Factor class represents each category with a number, but then
	## displays each number with the category label.  For this reason,
	## factors have both an as.numeric() and as.character()
	## interpretation.

head(factor_vector)
head( as.character(factor_vector) )
head( as.numeric(factor_vector) )

## Because it has both numeric and character properties, factor
	## class coercion results can be inconsistent across functions.
	## For example, unique() preserves the factor but intersect()
	## coerces to character.

unique(factor_vector)
intersect(factor_vector, factor_vector)

## Moreover, the relationship between categories numbers and
	## labels is fragile.  Essentially, each category number just
	## tells R which element of the levels() vector should be
	## displayed.

head(factor_vector)
as.numeric( head(factor_vector) )

levels(factor_vector) <- c("QQ", levels(factor_vector))
head(factor_vector)
as.numeric( head(factor_vector) )

## Delete demonstration objects that are no longer needed
remove( factor_vector )

```

Given these quirks, avoid factor class unless you need its unique properties.  A character vector will suffice for most data shaping purposes.

#### POSIXct Class - Managing Chronology

Performing calculations with times and dates is a complicated task.  Months have 28, 30, or 31 days.  Years have 365 or 366 days.  Minutes, days, and weeks are increments of 60, 24, and 7 increments, relative to the preceding unit.  There are 24 different time zones, and many time zones have different names in different countries.  Just over a third of countries observe daylight savings time, but some jurisdictions within them do not.  Most countries have adopted the Gregorian calendar as the official civil calendar, but not all.

Computers keep track of time by counting the number of seconds that have elapsed since a specific start date in the Coordinated Universal Time (UTC) time zone, which corresponds to time at the longitudinal prime meridian.  The start date can be operating system or even program specific.  For example, Excel starts counting at January 0 of 1900, SAS counts from January 1st of 1960, and MATLAB counts from January 0 of 1 BC.  Unix-based operating systems and many computer languages use a set of standards called POSIX time, which starts counting from January 1st of 1970.  R uses POSIX time.  For example, R would internally represent the first second of the year 2000 as 946,702,800, because 946.7 million seconds elapsed between the first second of January 1st of 1970 and the first second of January 1st, 2000 (UTC).

In addition, different regions use different notations to display dates.  In the United States, it common to write Month-Day-Year.  However, Year-Month-Day and Day-Month-Year are the most common formats around the world.  Within each region, there is significant variation in formatting, such as different separators (1/1/1970 vs 1.1.1970).  Programmers typically use the ISO 8601 standardized format, otherwise known as "unambiguous format."  In unambiguous format, the first second of POSIX time in Universal Coordinated Time is: 1970-01-01 00:00:00+00:00.  That is Year-Month-Day, then Hour:Minute:Second in 24-hour time, and finally, whatever amount of time would need to be added to UTC time to align it with the current time zone.  For example, 6pm on July 4th, 2000 in New York City (EST time zone, observing daylight saving) would be 2000-07-04 18:00:01-04:00.   It is acceptable to truncate this to just the date when applicable.  For example, 2000-07-04 would suffice to notate July 4th, 2000.
In short, chronology and chronological notation is perhaps one of the most complex, messy, vexing technical problems about which you have never needed to think.   As a programmer, you will become aware of its nuances.

```{R Chronology}
## The Sys.time() functions expresses the current time/date.
	## Note the use of unambiguous format.

Sys.time()

## Sys.timezone() expresses the current time zone localization
	## information.  This information can be useful as an input to
	## other chronological functions.  The listed city does not
	## correspond to the physical location of the programmer.  Instead
	## it is a reference city for that time zone.  The function 
	## OlsonNames() expresses a complete list of time zones that R
	## can interpret.

Sys.timezone()

head( OlsonNames() )

## as.POSIXct() is this lesson's recommended as._() function for chronological
	## coercion.  R supports multiple time standards, so it is important to
	## not confuse different date objects of different classes. To
	## convert a date to POSIXct class, enter a text string that 
	## records a date in unambiguous format.  Bear in mind that the
	## time of day is recorded in military time. The tz=
	## argument specifies time zone.  "UTC" serves as kind of a
	## world time zone, and correspond to the local time at the
	## Greenwich Meridian Observatory near London.

utc_date <- as.POSIXct( "2001-02-03 15:16:17", tz= "UTC" )
utc_date

## The underlying data for a chronological object is the number of
	## seconds that have elapsed since a time of origin.  That means
	## that POSIXct objects respond to arithmetic operations as if
	## doing calculations in units of seconds.

utc_date - 3600  ## go back an hour { 60 * 60 }
utc_date - 86400 ## go back a day { 60 * 60 * 24 }

as.numeric( utc_date )

## To find the origin date, subtract the numerical equivalent from
	## any POSIXct object.  This origin can be used for converting
	## seconds into meaningful time units.  In the example below,
	## the origin= argument, combined with a numeric measurement of
	## time is used to convert the origin time into UTC. When you
	## need to convert dates to a common time zone, this is one
	## approach for doing so.  This example uses "EST" which is a
	## shorthand abbreviation for the time zone containing the United States'
  ## East Coast.


est_origin <- Sys.time() - as.numeric(Sys.time())
est_origin

utc_origin <- as.POSIXct(
	x= as.numeric(est_origin),
	origin= est_origin,
	tz= "UTC"
)
utc_origin

## Remember that time zones are display characteristics of date
	## objects.  The underlying data (the number of seconds since the
	## origin) is the same.

est_date <- as.POSIXct(
	x= as.numeric(utc_date),
	origin= utc_origin,
	tz= "America/New_York"
)

est_date
utc_date

as.numeric(est_date)
as.numeric(utc_date)

## This applies to dates as well as times.  This example uses "JST"
	## which is an abbreviation for the time zone containing Tokyo.
	## While all three objects express the same moment in time (about
	## 981.2 million seconds since 1970), this moment occurred on the
	## 4th in Tokyo, and on the 3rd in New York and London.

jst_date <- as.POSIXct(
	x= as.numeric(utc_date),
	origin= utc_origin,
	tz= "Asia/Tokyo"
)

jst_date
utc_date
est_date

## The difftime() function measures differences in time. It is
	## important to specify the units= argument, as the default
	## for this function is to use a unit of time (seconds, hours,
	## weeks, etc) commensurate with the magnitude of difference
	## between dates. While convenient, this makes the function's
  ## results unstable unless you specify units=

since_origin_secs <- difftime(
	utc_date,
	est_origin,
	units= "secs"
	)

since_origin_hours <- difftime(
	utc_date,
	est_origin,
	units= "hours"
	)

since_origin_weeks <- difftime(
	utc_date,
	est_origin,
	units= "weeks"
	)

since_origin_secs
since_origin_hours
since_origin_weeks

## Objects emerging from difftime() are of class difftime.  While
	## they have an as.numeric() interpretation, it is no longer
	## seconds since origin.  However, you can use difftime objects
	## in calculations with POSIXct. R will convert the difftime
	## object to the correct unit when making the calculation.

since_origin_hours
as.numeric( since_origin_hours )

jst_date - since_origin_secs
jst_date - since_origin_hours

## Delete demonstration objects that are no longer needed
remove(est_date, est_origin, utc_date, utc_origin, jst_date,
	since_origin_hours, since_origin_secs, since_origin_weeks)
```

### Advanced Vector Indexing
In lesson A1, we explored how every data element in R has a position.  Using`[]`, all elements are retrievable by position.  The positioning system supplies abundant opportunities to use basic operations to achieve sophisticated object manipulations.  This lesson returns to positions now that you have learned more R programming.  

All vectors support the use of `names()` meta-data.  Positions can have designations called "names" that work like position number for retrieving elements:

```{R Position Names}
## create a vector of the first five letters
named_vector <- letters[1:5]

## Use names() to provide a designation for each position
names(named_vector) <- c( "Alpha", "Bravo", "Charlie", "Delta", "Echo" )
named_vector

## Names work in place of position numbers for retrieving elements
named_vector[c("Bravo", "Delta")]

## The data within a vector can be modified without altering the 
	## names.  This line of code uses empty brackets to select
	## all elements within the vector.  Without the empty brackets
	## the assignment operator would create a completely new object,
	## wiping out the name's meta-data in the process.  Also note how
	## coercion occurs in this example, since the object is of
	## mode character.

named_vector[] <- 1:5
named_vector

named_vector <- 1:5
named_vector

## Operations performed element-wise on an object generally do not influence the
  ## names meta-data.  However, many operations will wipe out names.  When
  ## programming, you will need to decide whether the risks of using names
  ## (i.e., that a function wipes them out) if worth the benefit of names
  ## (i.e., lower chances of programmer error when selecting within vectors)

names(named_vector) <- c("Alpha", "Bravo", "Charlie", "Delta", "Echo")

named_vector / 2
cumsum(named_vector)
diff(named_vector)

sum(named_vector)
as.character(named_vector)

## Names do not have to be unique.  This example uses append() to
	## add an element to named_vector that is also named Delta. The after=
	## argument specifies the position after which the new elements will
  ## be inserted.  With the new element inserted, there are now two
  ## elements with the same name.  However, R will only retrieve
	## the element with the name Delta that has the lowest position
	## number.

named_vector <- append(
	x= named_vector,
	values= c("Delta"= 33),
	after= 3
	)

named_vector
named_vector["Delta"]

## To retrieve all elements named Delta, use a logical test of names.
	## The %in% operator works for situation when multiple names
	## are applicable.

named_vector[names(named_vector) == "Delta"]
named_vector[names(named_vector) %in% c("Charlie", "Delta")]

## As your indexing operations become more complex, you may want to
	## separate the logical testing step from the indexing step
	## and/or store the result of the logical testing step. which()
	## stores those results in a parsimonious format.  As you begin
	## to work with objects at scale, this difference in resource
	## use between a logical vector, and its which() representation
	## becomes useful for efficient computation.

index <- names(named_vector) %in% c("Charlie", "Delta")
index

index <- which(index)
index

named_vector[index]

## Delete demonstration objects that are no longer needed
remove(index, named_vector)
```

R has many functions that operate primarily via position manipulations.  For example, there are many instances where it is useful to either put a collection of things in a specific order, or to reorder one collection of things to align with the order of another collection.  The functions sort() and order() will re-order elements to match alphanumeric order.

```{R Position Manipulations with order() and sort()}
## create an object that contains states by population and shuffle
state_pop <- c(
	"CA"= 40, "TX"= 29, "FL"= 21, "NY"= 20, "PA"= 13,
	"IL"= 13, "OH"= 12, "GA"= 11, "NC"= 10, "MI"= 10
	)

set.seed(3333)
state_pop <- sample(state_pop)

#### ---- Re-sort Elements into Alphanumeric Order 

## The sort() function sorts the elements of a vector by alphanumeric
	## order.  Here sort() is used to first sort by population size,
	## ie by the elements of the vector, and then used to sort by 
	## alphabetical order, ie by the elements of the names() vector.

sort(state_pop)

state_pop[sort(names(state_pop))]

## The sort() function is actually a convenience function derived from
	## order().  order() returns the positions of a vector, sorted
	## according to alphanumeric order.

pop_order <- order(state_pop)
pop_order
state_pop[pop_order]

alpha_order <- order( names(state_pop) )
alpha_order
state_pop[alpha_order]

## Delete demonstration objects that are no longer needed
remove(pop_order, alpha_order, state_pop)
```

Using match() in tandem with order(), objects can be sorted into any arbitrary order.  The demonstration below exemplifies different approaches to achieving a desired order.

```{R Position Manipulation with match() and order()}
## Create an object that contains states by population and shuffle
state_pop <- c(
	"CA"= 40, "TX"= 29, "FL"= 21, "NY"= 20, "PA"= 13,
	"IL"= 13, "OH"= 12, "GA"= 11, "NC"= 10, "MI"= 10
	)

set.seed(3333)
state_pop <- sample(state_pop)

## Create an object that contains state names and abbreviations
state_abbrev <- c(
	"California"= "CA", "Texas"= "TX", "Florida"= "FL",
	"New York"= "NY", "Pennsylvania"= "PA", "Illinois"= "IL",
	"Ohio"="OH", "Georgia"= "GA", "North Carolina"= "NC",
	"Michigan"= "MI", "New Jersey"= "NJ", "Virginia"= "VA"
	)
set.seed(3120)
state_abbrev <- sample(state_abbrev)

#### ---- Different Approaches to Aligning Order

## Aligning two vectors by using names.  Note how the difference
	## between the 10 element state_pop and 12 element state_abbrev
	## is resolved using the %in% operator.

index <- state_abbrev[ state_abbrev %in% names(state_pop) ]

state_pop[index]
state_abbrev

## Aligning two vectors to an arbitrary order using match():  For this
	## example, the goal is align both vectors to alphabetical order
	## of the full state names.  For each element of the first 
	## vector, match() will tell you the position of a matching
	## element in the second vector.

## Create a vector that has the desired ordering.
desired_order <- sort( names(state_abbrev) )
desired_order

## Identify where each element of a vector falls in the desired order
index_abbrev <- match(names(state_abbrev), desired_order)

index_abbrev

state_abbrev
desired_order[index_abbrev]

## Re-order those positions to alphanumeric order to match the desired

desired_order
state_abbrev[order(index_abbrev)]

## Now state_pop can be reordered, despite not containing full names
index_pop <- state_abbrev[order(index_abbrev)]
index_pop <- match(names(state_pop), index_pop)


desired_order
state_abbrev[order(index_abbrev)]
state_pop[order(index_pop)]

## Delete demonstration objects that are no longer needed
remove(state_pop, index_pop, index_abbrev, desired_order, state_abbrev, index)
```

For completeness, here is another example of match() and order() used in tandem.

```{R  Second example with match() and order()}
## Generate variables
shirt_sizes <- c( "S", "M", "L", "XL" )

set.seed(5043)
shirt_orders <- sample(
	shirt_sizes,
	size= 2^4,
	replace= TRUE
	)

shirt_sizes
shirt_orders

## order() returns a vector of position that, if supplied to square
	## brackets, would place a vector of objects in alphanumeric
	## order.

shirt_sizes[order(shirt_sizes)]
shirt_orders[order(shirt_orders)]

## For each element in the first vector, match() returns the position
	## of a corresponding element in the second vector.  If supplied
	## to square brackets on the second vector, it can be used to
	## generate an exact duplicate of the first vector.

shirt_orders
shirt_sizes[match(shirt_orders, shirt_sizes)]

## Using match() and order() in tandem, the elements of the first
	## vector can be re-ordered to match the ordering of the second

shirt_sizes
shirt_orders[order( match(shirt_orders, shirt_sizes) )]

## Delete demonstration objects that are no longer needed
remove(shirt_sizes, shirt_orders)
```

### Functions

In the Object Inspection section of this lesson, we discussed the importance of devising tests to ensure your objects have the properties you think they have.  For a particular use case, you may need to run multiple, complex tests at intervals to ensure that your code is working as intended.  This is an example of a common occurrence in programming – you will frequently need to run entire sections of code multiple times, either because the same transformation needs to be applied to multiple objects, or because the same tests need to be performed after multiple steps.  Any time you find yourself wanted to copy/paste code, it is time to build your own custom function to hold that code instead.  `The function(){}` function enables you to write your own functions.  Below is an example of `function(){}` being used to package multiple tests into one convenient function.

The object names between the `()` specify the arguments that the function will expect to receive.  The `=` operator enables you to specify the default for that argument.  If no default is specified, the function will throw an error if the user does not supply an object for that argument when executing the function.  In this example `fix_negatives=` has a default, but `numbers=` does not.  The code between the `{}` is the code that will be applied to the objects specified as arguments.  If you had not written a custom function, this is the code you would need to copy and paste.  The last line of code in a function is typically an expression.  This specifies what the function will express once it has completed its calculations.  Some style guides recommend encasing this expression in a `return()` function and others do not.  For the duration of this class, please do.

```{R Diagnostic Functions}
## Create an object suitable for testing
normal_numbers <- rnorm(n= 10^4, mean= 100, sd= 16)

DiagnosticTest <- function(numbers, fix_negatives= TRUE) {

## Create a blank vector to hold outputs
diagnostics <- vector(length= 4)
names(diagnostics) <- c("Positive Numbers", "Numeric Class", "No NAs",
	"Normal Tails")

## Get rid of negative numbers if fix_negatives == TRUE
neg_index <- numbers < 0
diagnostics["Positive Numbers"]  <- !any(neg_index)
numbers[neg_index] <- numbers[neg_index] * !fix_negatives

## Test that object is of the correct class
diagnostics["Numeric Class"] <- is.numeric(numbers)

## Test that there are no NAs in the object
diagnostics["No NAs"] <- !any( is.na(numbers) )

## Test that tails contain one-third of probability mass
numbers_1sd <- mean(numbers) + {sd(numbers) * c(-1, 1)}
names(numbers_1sd) <- c("Lower", "Upper")

tail_mass <- {numbers >= numbers_1sd[ "Upper" ]} | {numbers <= numbers_1sd[ "Lower" ]}
tail_mass <- mean(tail_mass)

diagnostics["Normal Tails"] <- abs(tail_mass - 0.333) <= 0.05

return(diagnostics)
}

DiagnosticTest(normal_numbers)

## Delete objects that are no longer needed
remove(normal_numbers, DiagnosticTest)
```

This example demonstrates an object testing function.   As discussed above, a programmer might express it at intervals to confirm that code is running as intended.  However, these types of function also often appear as the first line of a complex analytic function to double check that the user has correctly specified the arguments for that function.

### Statistical Distributions

Not only can R generate random numbers in different statistical distributions, it also has a full suite of functions for conducting analysis of statistical distributions.  This section uses t-distribution statistics to show these capabilities.

```{R Statistical Distributions}
## the r_() functions generate random numbers from a given statistical
  ## distribution.

set.seed(4602)
random_tdist <- rt(n= 10^6, df= 10^9)

round(summary(random_tdist), 2)

## the p_() statistical distribution functions accept a number, and
  ## calculate the chances that a number less than or equal to that
  ## number would generate from that distribution.

percentiles_tdist <- pt(q= -2:2, df= 10^9)

round(percentiles_tdist, 2)

## the q_() statistical distribution function accepts a quantile,
  ## and returns the number.  p_() and q_() functions are inverses
  ## of each other, converting between quantiles of a distribution
  ## and the numbers in that distribution

qt(p= percentiles_tdist, df= 10^9)

qt( pt(q= -2:2, df= 10^9), df= 10^9)

## the d_() statistical distribution function accepts a number, and
  ## returns the percentage of random numbers drawn from that
  ## distribution that would equal that number.  If you fed
  ## dt() outputs into the prob= argument for sample(), you could
  ## use sample() to mimic the outputs of rt()

dt(-1, df= 10^9)

mean(round(random_tdist)  == -1)

set.seed(1014)
simulated_rt <- sample(
  x= seq(from= -5, to= 5, by= 0.01),
  prob= dt( seq(from= -5, to= 5, by= 0.01), df= 10^9),
  size= 10^6,
  replace= TRUE
)

round( summary(simulated_rt), 2)
round( summary(random_tdist), 2)

## Delete demonstration objects that are no longer needed
remove(simulated_rt, random_tdist, percentiles_tdist)
```


Building on the t-distribution functions, the `t.test()` function provides the classic statistical tests for differences in means. Like many statistical test functions in R, `t.test()` supports two methods of argumentation.  The first method supplies `x=` and `y=` arguments, each of which is a numeric vector representing one of the comparison samples.  The second method supplies symbolic instructions to the `formula=` argument.  In R, the `~` operator indicates that the arguments supplied constitute a formula, with something like a dependent variable on the left and something like an independent variable on the right.  For t-tests, a numeric vector of measurements goes on the left of the `~` and a factor-class vector goes on the right.  That factor identifies to which sample population that measurement belongs. 

```{R T-Tests}
## Generate normally distributed heights (based on CDC data)
set.seed(3054)
height_female <- rnorm(mean= 5.357, sd= 0.212, n= 10^3)
height_male <- rnorm(mean= 5.800, sd= 0.234, n= 10^3)
height_measurements <- c(height_female, height_male)

gender_factor <- as.factor( rep( c( "F", "M" ), each= 10^3) )

## Test for difference in means using the x= , y= arguments
t.test(x= height_female, y= height_male)

## Test for difference in means using the formula= argument
t.test( formula= height_measurements~gender_factor )

## Delete objects that are no longer needed
remove(height_female, height_male, height_measurements, gender_factor)
```

Like the factor and POSIXct classes, t.test() relies on the display instructions for its class to generate sophisticated expressed output from simpler underlying data.  We will explore objects like `t.test()` in more detail in Lesson B1.

The `t.test()` function illustrates a basic principle of programming – all complex programs are built up from simple functions.  To get complex programs, a programmer combines simple functions into sequences that cumulatively generate sophisticated results.  At its core, `t.test()` is an interface for using t-distribution functions to test for statistical significance.  Below is a table of the functions it uses to accomplish this result, how many times each is used, and whether the function is covered is this textbook.  Of the 263 functions and operators in `t.test()`, 98 percent are basic functions covered in the vocabulary.

|Function		|Count		|Lesson |Function		|Count		|Lesson
|:-					|:-				|:-		|:-					|:-				|:-
|`<-`				|53		|A1		|`if(){}`		|22		|B5
|`-`				|19		|A1		|`=`				|13		|A2
|`!`				|13		|A1		|`/`				|12		|A1
|`<`				|11		|A1		|`+`				|10		|A1
|`else{}`		|9		|B5		|`stop()`		|9		|B5
|`*`				|9		|A1		|`|`				|7		|A1
|`c()`			|6		|A2		|`&`				|5		|A1
|`^`				|5		|A1		|`sqrt()`		|5		|A2
|`abs()`		|4		|A2		|`>`				|4		|A1
|`is.na()`	|4		|A3		|`length()`	|4		|A2
|`names()`	|4		|A4		|`deparse1()`|3		|
|`is.null()`|3		|A3		|`pt()`			|3		|A4
|`qt()`			|3		|A4		|`substitute()`|3	|
|`==`				|2		|A1		|`mean()`		|2		|A2
|`missing()`|2		|X		|`paste()`	|2		|A5
|`var()`		|2		|A2 teaches closely related `sd()`|`attr()`		|1		|B5
|`class()`	|1		|A3		|`complete.cases()`|1		|
|`$`							|1		|B1		|`function(){}`		|1		|A2
|`is.finite()`		|1		|A3 teaches closely related `is.infinite()`|`list()`					|1		|B1
|`match.arg()`		|1		||`max()`					|1		|A2
|`setNames()`			|1		|A4 teaches closely related `names()`|||

```{R Anatomy of t.test}
## If you want to see the source code for t.test (x=, y= method), you can use
  ## this expression:

# getAnywhere( methods(t.test)[1] )
```

### Vocabulary Table for Lesson A4

In order to program effectively, you will need to memorize basic functions, operators, and constants.  Write each of the functions/operators/constants below on a flash card.  On the back of each card, write a succinct definition of what it does and a example of a line code you could enter into console that uses it.  Drill with these cards until you have memorized them.  Then drill again, coming up with a fresh example for each and testing that example in the console.

In order to understand what each function/operator/constant does, use the `help()` function to pull the documentation for it.  For example, `help("objects")` would pull up the documentation for the function `objects()`.  This document includes a description of what the function does ("Description" section), a list of all the arguments that can be given to the function ("Arguments" section), and examples of how to use the function ("Examples" section) at the bottom.  Only copy the definition or example from the documentation to your flash card if you absolutely understand what it does.  Otherwise, substitute your own.

The help documentation may be a difficult to read at first but keep practicing.  Over time, getting useful information from the documentation will become effortless.  Resist the impulse to do a Google search before you have consulted the documentation.  Google results can be of mixed quality - sometimes you will get a thoughtful, efficient solution, sometimes you will get a byzantine work-around that will teach you bad habits.

|Inspecting Objects		|Factors		|Chronology		|Indexing		|Distributions
|:-						|:-							|:-								|:-					|:-
|`tail()`				|`cut()`					|`as.POSIXct()`			|`append()`		|`~`
|`head()`				|`factor()`				|`difftime()`				|`match()`		|`dt()`
|`quantile()`		|`levels()`				|`OlsonNames()`			|`names()`		|`pt()`
|`summary()`		|`nlevels()`			|`Sys.time()`				|`order()`		|`qt()`
|							|								|`Sys.timezone()`		|`sort()`			|`rt()`
|							|								|`weekdays()`				|`which()`		|`t.test()`
