```{r Environment set-up, include= FALSE}
remove(list= objects())
```

```{r Document-Wide Knitr Options, include= FALSE }
knitr::opts_chunk$set(fig.width=6, fig.height=6) 
knitr::opts_chunk$set(comment= "")
knitr::opts_chunk$set(prompt= TRUE)
```
## Lesson Links

[Go to the Table of Contents](00_TableOfContents.html)

[Return to Last Lesson (B1: Array and List Objects)](06_LessonB1.html)

[Proceed to Next Lesson (B3: Environment)](08_LessonB3.html)

## B2.  Segmentation

==========
==========

### Vocabulary Table for Lesson B2

In order to program effectively, you will need to memorize basic functions, operators, and constants.  Write each of the functions/operators/constants below on a flash card.  On the back of each card, write a succinct definition of what it does and a example of a line code you could enter into console that uses it.  Drill with these cards until you have memorized them.  Then drill again, coming up with a fresh example for each and testing that example in the console.

In order to understand what each function/operator/constant does, use the `help()` function to pull the documentation for it.  For example, `help("objects")` would pull up the documentation for the function `objects()`.  This document includes a description of what the function does ("Description" section), a list of all the arguments that can be given to the function ("Arguments" section), and examples of how to use the function ("Examples" section) at the bottom.  Only copy the definition or example from the documentation to your flash card if you absolutely understand what it does.  Otherwise, substitute your own.

The help documentation may be a difficult to read at first but keep practicing.  Over time, getting useful information from the documentation will become effortless.  Resist the impulse to do a Google search before you have consulted the documentation.  Google results can be of mixed quality - sometimes you will get a thoughtful, efficient solution, sometimes you will get a byzantine work-around that will teach you bad habits.

|Importing Libraries	|Basic Apply Family		|Adaptive Apply-Like Functions		| Parallel Processing (package:parallel)
|:-										|:-					|:-									|:-		
|`data()`								|`apply()`		|`do.call()`					|`detectCores()`
|`detach()`							|`lapply()`		|`eapply()`						|`makeCluster()`
|`install.packages()`		|`mapply()`		|`environment()`			|`parApply()`
|`library()`						|`sapply()`		|`rapply()`						|`parLapply()`
|`search()`							|`tapply()`		|`simplify2array()`		|`stopCluster()`
|`update.packages()`												
