---
  title: B3.  Environment
---

```{r Environment set-up, include= FALSE}
remove(list= objects())
```

```{r Document-Wide Knitr Options, include= FALSE }
knitr::opts_chunk$set(fig.width=6, fig.height=6) 
knitr::opts_chunk$set(comment= "")
knitr::opts_chunk$set(prompt= TRUE)
```

[Go to the Table of Contents](00_TableOfContents.html)

[Return to Last Lesson (B2: Segmentation)](07_LessonB2.html)

[Proceed to Next Lesson (B4: Visualization)](09_LessonB4.html)

This lesson explores the ways that R can interact with the digital world around it.  This includes reading files from and writing them to your computer, directly manipulating files on your computer, retrieving files (including webpages) from the Internet, executing other R scripts, and sending instruction to your operating system.

### Think Before You Execute

In this lesson, you will learn how to bridge the boundaries between the R environment and the digital world around it.  Before this lesson, there was nothing you could do using the class material that a reboot would not fix.  With this lesson, that ceases to be true.  An incorrect line of code can potential corrupt your data or damage files critical to your operating system.  Think before you execute.

### Read and Writing to Memory

In a computer file system, every file has a file path.  This file path provides complete directions for finding a file within hierarchical layers of directories.  For example, on a Mac, the desktop is a directory located inside your user account directory which is located inside the User directory which is a top-level directory of the file system.  So, the full file path for a file on your desktop would be Users => UserName => Desktop => YourFile.ext.  In Mac (i.e., Unix-like) file path syntax, this would be denoted as `/Users/UserName/Desktop/YourFile.ext`.  Even though you only see the `YourFile.ext` part when viewing the file through your operating system's graphical interface, the full file path is the actual designation of the file.  That is why you can have files in different directories that have the same name.  On a PC, file path syntax is similar.  However, PCs use backslashes instead of forward slashes.  Since backslash is the escape character in R, double backslash denotes a literal backslash.  So, the file path shown above would be `\\Users\\UserName\\Desktop\\YourFile.ext`.   In addition, PCs use a different way of handling volumes in path names, so a PC-style file path is likely to have a volume letter, followed by a colon, as in `C:\\Users\\UserName\\Desktop\\YourFile.ext`.  As present, forward slashes work on both Unix-like and Windows-based operating systems, so they are the recommended way to specify file paths.

When you open R, the program opens with an implicit location within the file directories of your system.  This location is the "working directory".  The `getwd()` function will report your current working directory.  The `setwd()` function enables you to change your working directory.  When you read or write files and only specify the file name, R will assume that you want to access a file by that name that is situated in your working directory.  A full file path that has the directory lineage all the way back to top-level is an "absolute" file path.  A file path that only denotes where a file should be relative to the working directory is called a "relative" file path.

```{R Working Directory}
## Determine working directory
getwd()

## Set working directory
setwd("A_Inputs")
getwd()

## Use .. to go one level back up in the file directory
setwd("..")
getwd()
```

The `save()` function will save R objects to your file system in R's native storage format.  The `load()` function will read R objects from files in that format and make those objects available in your environment.  While `save()` and `load()` do not depend on file extensions to perform their functions, general practice is to give your save files the extension of `.RData` when saving.

```{R save() and load()}
## Create an object to save
save_object <- as.list(letters[1:4])

## Save object to B_Intermediates as per Lesson A1

save(save_object, file= "B_Intermediates/ListObject.RData")

## Clear object from the environment
remove(save_object)
objects()

## Restore object from saved file
load( "B_Intermediates/ListObject.Rdata" )
objects()

## Delete demonstration object
remove(save_object)
file.remove("B_Intermediates/ListObject.Rdata")
```

The `load()` function violates the "expression versus assignment" behavior standard in R.  When you assign the expression from `load()` to an object, that object will only receive a character vector containing the names of the objects that `load()` added to the environment -- not the objects themselves.  You can use functions like `assign()`, `get()`, and `remove()` to manipulate objects using their names.

Aside from R's native file format, R supports plain text file formats.  This includes the basic `.txt` text file, and `.csv` spreadsheet file formats.  The `.txt` file is a simple file format that holds text characters and provides no other information (such as formatting).  The `.csv` also holds simple text characters.  However, it puts each row of a spreadsheet on its own line and uses a symbol called a "delimiter" to delineate the boundaries between cells.  This delimiter is typically a comma.  The example below displays a two x two spreadsheet, and the `.csv` encoding that would represent it.  Note how quotations distinguish between commas that are part of the cell contents and commas that are cell delimiters.

**Example Table**

|					|Column 1		|Column 2
|:-|:-|:-
|Row 1		|A					|B
|Row 2		|C,D				|E

**CSV Notation for the Table**

`,Column 1,Column 2\nRow 1,A,B\nRow2,"C,D"\n`

In R, functions that write data to the file system typically follow the naming pattern of `write._()` and functions that read data from the file system typically follow the naming pattern of `read._()`.  For example, the file functions for `.csv` files are `read.csv()` and `write.csv()`.  However, the .txt file function breaks this pattern, using `readLines()` and `writeLines()`.  It also uses `con=` instead of the standard `file=` for the argument specifying the file path.  

The `writeLines()` function interprets each element of a `text=`character vector as its own line in the resulting text file.  However, any `\n` within those elements will also be treated as denoting a new line.  The `write.csv()` function looks for an `x=` array or data frame argument.  

If you do not specify a `file=` argument for `write.csv()` or a `con=` argument for `writeLines()`, the function will express (in the console) what it would have written to the file system.

```{R txt and csv expressed to console}
## write.txt file output, expressed in console
writeLines(text= c("A", "AB", "ABC", "ABCD\nEF"))


## write.csv file output, expressed in console
write.csv(
	x= array(1:4,
		dim= c(2, 2),
		dimnames= list("Row"= letters[1:2], "Col"= LETTERS[1:2])
		)
	)
```
For `read.csv()`, note the `header=` argument.  When set to `TRUE`, `read.csv()` will treat the first line of the file as the column names for the data frame it generates.

### File Formats and Translation Packages

`foreign`– Like `parallel`, the `foreign` package comes pre-installed in basic R but you must load it for each session using the `library()` function.  It supplies functions for reading data files in the formats of some statistical analysis programs.   This includes limited support for SAS, SPSS, Octave, Minitab, Microsoft Access, and older versions of STATA files.  These functions tend to follow the standard function and argument naming conventions for read/write functions.  However, there are at least three major gaps in `foreign`, for which you may need to install other packages.

1. `openxlsx`– `foreign` does not have a function for reading Microsoft Excel `.xlsx` files.  I recommend installing package `openxlsx`.  Avoid package `xlsx` as it depends on Java code calls that are vulnerable to performance issues.

2. `readstata13`– `foreign` only reads version 5-12 STATA files.  The `foreign` package team recommends `readstata13`.  Note: STATA supports multiple missing-ness codes, while R only represents missing-ness with NA, so some information may be lost in translation.  

3. `rjson`– `foreign` does not currently have a function for reading `.json` formatted data. This format is common for data stored on webpages or transmitted through server Application Programming Interfaces (API).  The `.json` format is short for JavaScript Object Notation.  Its syntax is based on JavaScript, the programming language underlying most of the interactive elements of modern websites.  In form, a `.json` objects resembles an R list object with data frams objects or vectors contained within.  The data structure was designed to be both human-readable and machine-readable.  The package `rjson` is a good option, if not the only option, for reading/writing `.json` files.

### File System Manipulation

R can create, move, copy, rename, and delete files and directories from you file system.  The `list.files()` function is analogous to `objects()`.  It expresses the file paths for the files in your working directory.  With the `include.dirs=` argument set to `TRUE`, the expression will include sub-directories within your working directory.  With the `recursive=` argument set to `TRUE`, the expression will include all files in sub-directories within your working directory.

```{R File System Manipulation}
## Set working directory
# setwd("/Users/smendels/Desktop/ExampleProject")

## View contents of current working directory
list.files(include.dirs= TRUE)

## Confirm whether standard files and directories exist
dir.exists( c("A_Inputs", "B_Intermediates", "C_Outputs"))
file.exists("README.md")

## create files for file manipulation demo
writeLines(
	text= "To be, or not to be, that is the question:\nWhether 'tis",
	con= "B_Intermediates/Hamlet.txt"
	)
write.csv(
	x= array(1:6, dim= c(3, 2), dimnames= list(letters[1:3], LETTERS[1:2])),
	file= "B_Intermediates/Numbers.csv"
	)

## View contents of working directory and all sub-directories,
	## confirming that new files now exist.  This function ignores all of the 
  ## textbook files, which would make the outputs unwieldy.
ListFilesIgnoreLessons <- function(){
	file_list <- list.files(recursive= TRUE, include.dirs= TRUE)
	file_list[-grep("[.](html|Rmd)$", file_list)]
	}
ListFilesIgnoreLessons()

## Move file to new directory and rename
file.copy(from= "B_Intermediates/Hamlet.txt", to= "C_Outputs/Words.txt")
ListFilesIgnoreLessons()

## Delete all files with a given extension, using file.remove()
	## and list.files() in tandem.
txt_list <- list.files(pattern= "[.]txt$", recursive= TRUE)
txt_list <- txt_list[!grepl("A_Inputs", txt_list)]
txt_list

file.remove(txt_list)
ListFilesIgnoreLessons()

## Create a new directory
dir.create("E_NewDirectory")
ListFilesIgnoreLessons()

dir.create("E_NewDirectory") ## note warning if dir. already exists

## Delete remaining demonstration files
file.remove( c("B_Intermediates/Numbers.csv", "E_NewDirectory"))
remove(ListFilesIgnoreLessons, txt_list)
```

These functions can be especially powerful (and hazardous) in combination with segmented processing functions.  Here is an example of how to use them in conjunction to do a file processing task at scale.  This script first generates 1,000 csv files, then reads them back in, and finally compiles them to a single dataset.  On my machine, this takes about 5 seconds.

```{R File Processing at Scale}
## Load the parallel package
library(parallel)

## Generate 1,000 xlsx files ==========

## Create folder
dir.create("B_Intermediates/SourceFiles", showWarnings= FALSE)

## Create function to generate files
GenerateFiles <- function(seed_number) {

	## set PRNG seed
	set.seed(seed_number)

	## Generate dataset
	new.data <- data.frame(
		"Type"= sample(x= letters, size= 100, replace= TRUE),
		"Score1"= rpois(n= 100, lambda= 3),
		"Score2"= rpois(n= 100, lambda= 3)
		)

	## Write dataset
	write.csv(
		x= new.data,
		file= paste("B_Intermediates/SourceFiles/", seed_number, ".csv", sep= "" ),
		row.names= FALSE
		)
	}


##  Generate and write files
  ## The output_bucket diverts the expressions from parLapply() to a harmless
  ## container object, so that it does not flood the console.  The function
  ## invisible() would have accomplished the same purpose.
core_cluster <- makeCluster( floor(detectCores() * 0.8))

output_bucket <- parLapply(
	cl= core_cluster,
	X= as.list(1000:1999),
	fun= GenerateFiles
	)

stopCluster(core_cluster)

## Collect all datasets and compile ==========

## Create roster of data files to import
data_roster <- list.files(
	path= "B_Intermediates/SourceFiles",
	pattern= "[.]csv$",
	full.names= TRUE
	)
head(data_roster)

## read in all files
core_cluster <- makeCluster( floor(detectCores() * 0.8))

all_data <- parLapply(
	cl= core_cluster,
	X= as.list(data_roster),
	fun= read.csv
	)

stopCluster(core_cluster)

lapply(all_data[1:2], head)

## Compile to single dataset, noting file of origin for each row
file_names <- rep(
	x= data_roster,
	times= sapply(all_data, function(x) {dim(x)[1]})
)

all_data <- do.call(
	what= rbind,
	args= all_data
	)
all_data <- data.frame("Files"= file_names, all_data)

head(all_data)
tail(all_data)
dim(all_data)

## Rename source file directory once compilation is done (prevents
	## data contamination from accidental second run, will cause
	## harmless error if second run occurs without deleting this
	## directory).  We'll discuss more sophisticated options for
	## this in B5.
list.files("B_Intermediates", include.dirs= TRUE)

file.rename(
	from= "B_Intermediates/SourceFiles",
	to= "B_Intermediates/COMPILEDSourceFiles"
	)

list.files("B_Intermediates", include.dirs= TRUE)

## Delete demonstration objects and files; reset packages
detach(package:parallel)
output_bucket <- file.remove(
	list.files("B_Intermediates", recursive= TRUE, full.names= TRUE))
remove(all_data, core_cluster, data_roster, GenerateFiles, output_bucket,
	file_names)
file.remove("B_Intermediates/COMPILEDSourceFiles")
```

### Internet File Retrieval

For R, the distinction between digital resources on your computer and digital resources on the Internet more minor than you might think.  Like local files, Internet pages are files and have file paths (URLs) – typically `.html` files.  As such, you can point most file-reading function at the web.  In the example below, `readLines()` extracts the html code file from `https://www.example.com` , which is one of the official example and testing websites of the web.

```{R Accessing Web Resources}
## readLines() will read webpages effortlessly
example_html <- readLines(con= "https://www.example.com")
example_html

## Delete demonstration object
remove(example_html)
```

The `url()` and `close()` functions provide formal management of connections between your computer and the Internet.  The former creates a connections object you can supply to functions, and the latter closes that connection.

```{R url()}
## Create a connection
example_connect <- url("https://www.example.com")
example_connect
class(example_connect)

## Retrieve webpage
example_html <- readLines(con= example_connect)

## Close the connection
close(example_connect)
example_connect
class(example_connect)

## Delete demonstration objects
remove(example_connect, example_html)
```

`readLines()` can read web pages because `.html` format is plain text with special meaning, just like `.csv` and `.json` files.  For files in a less accessible format, `download.file()` can download a file without trying to read it.

File compression makes a file occupy less space in your system's file system (i.e., storage memory) by using a kind of shorthand to represent the data it holds.  This makes the file inaccessible to any program that cannot decode that shorthand and slower to load for programs that can.  However, it is useful for situations where you need to archive a large amount of data.  It is also useful for doing Internet file transfers.  Compressed files, by definition, are smaller and that translates to less download time.  The `.zip` family of algorithms is the most ubiquitous form of file compression.  R can compress with `zip()` and decompress with `unzip()`.  To measure the effectiveness of compression, the `object.size()` and `file.size()` functions measure the size of the object in memory and in the file system respectively.  In the example below, note how the data is over twice as large as a .csv file, compared to its RAM size, and four times as large as its compressed '.zip' form.  Also note that R applies compression automatically to the objects it saves in `.rdata` internal format.

```{R File Compression}
## create large array of numbers
large_array <- array(1:10^8, dim= c(10^6, 10^2))

## measure the size of the object in RAM
object.size(large_array)

round( unclass( object.size(large_array)) / 10^6, 2) ## in Mb

## write to disk and measure file size
write.csv(large_array, file= "C_Outputs/ManyNumbers.csv", row.names= FALSE)
list.files("C_Outputs")
round( file.size("C_Outputs/ManyNumbers.csv") / 10^6, 2)


## compress file and read file file
zip(files= "C_Outputs/ManyNumbers.csv", zipfile= "C_Outputs/ManyNumbers.zip")
round( file.size("C_Outputs/ManyNumbers.zip") / 10^6, 2)

## Delete demonstration objects and files
remove(large_array)
file.remove("C_Outputs/ManyNumbers.csv", "C_Outputs/ManyNumbers.zip")
```

### Interface Access
 
In complex coding projects, there are circumstances under which it is useful to let your code use decision rules to find a needed object, rather than hard coding the name of that object.  For example, if you are loading batches of `.RData` files from an external source, you may need an approach for identifying the names of the newly loaded objects and sorting / re-packaging them to fit an organizational schema.  The `get()` function retrieves an object based on its name, and the `assign()` function copies the contents of an object to another object  Used in tandem with `objects()` and `eapply()`, they supply many options for handling arbitrary collections of objects.

```{R Object Retrieval}
## Create three objects of lengths that randomly vary each run
RandomVector <- function(){
	rpois(
		n= rpois(n= 1, lambda= 10),
		lambda= 3
		)
	}

object.a <- RandomVector()
object.b <- RandomVector()
object.c <- RandomVector()

remove(RandomVector)

object.a

## determine which object is longest
long.vector <- environment()
long.vector <- eapply(env= long.vector, FUN= length)
long.vector <- unlist(long.vector)
long.vector <- long.vector[long.vector == max( long.vector )]
long.vector <- names(long.vector)[1]
long.vector

## create new object with "longest" in title to hold the longest vector
assign(x= paste("longest", long.vector, sep= "."), value= get(x= long.vector))
remove(list= c(long.vector, "long.vector"))

objects()

## Delete demonstration objects
remove( list= objects()[grepl("object.[abc]", objects())])
```

Thinking more broadly about the R processing environment, the `options()` and `source()` functions provide ways to make significant changes to R's behavior. The `options()` function supplies access to R's settings.  Here are some options to keep in mind:

* Change how R displays decimals to make terminal output easier to read.
  - `digits=` is a suggestion passed to functions on the number of significant digits for displaying decimals.  The default is seven, but you might find 4 easier to read.  For example, at `digits= 7`,` pi` displays as `3.141593`, but at `digits= 4`, it displays as `3.142`.
  - `scipen=` biases R's decision-making on when to start using scientific notation to display large / small numbers.  Positive numbers bias R away from notation, while negative numbers bias it towards notation.  Each unit corresponds to about 1 decimal place of bias.  For example, the default `scipen= 0` starts to use notation at `10^-4`, but `scipen= 1` starts to use notation at `10^-5`.  There may be merit to `scipen= 2`, which prevents notation from kicking in until the decimal is down to the millionths, or the number is up to 10 million.

* Change how R displays prompting characters if necessary.  However, there are very few circumstances that call for this.
  - `prompt=` sets the default terminal prompt string to `> `.
  - `continue=` sets the default continuation prompt string to `+ `.
  
*	Change how R displays high-volume text.
  -	`max.print=` Sets the total number of lines that R will echo in terminal before truncating the output.  The default (`99999`) is enormous.  Something like `999` will ensure that you cannot accidental flood your terminal with a torrent of output.
  - `width=` sets the maximum number of characters to display on each line.  For tidy code and readable terminal output, the longest line of your code should be no longer than this number and the width of your terminal window should be no shorter.

*	Change R's default behavior for specific types of activity
  -	`repos=` defines which CRAN repository R will use to download / update packages.
  -	`na.action=` specifies how R handles dataset cases with NA's when executing statistical procedures.  The default function, `na.omit()`, drops cases with `NA` before proceeding.  Alternatives might include` na.fail()` which will throw an error or `na.exclude()`, which pads residuals and predicted values back to the right length, inserting `NA`s is the right places.
  -	`error=` changes how R responds to non-catastrophic errors.  Setting this argument to the `recover()` function provides advanced tools for investigating errors but takes practice to master.

The `source()` function executes another R script.  This can be useful when working on projects that require many customized functions and system defaults.  It can also supply a short cut for running a large coding project from end to end to generate the final set of deliverables.  For example, you could have a script that looks like the one below, and results in the tables and figures you need being deposited in your project's `C_Outputs` directory.  The `echo= FALSE` argument prevents R from printing all of the code from those scripts in the console.

```{R source(), eval=FALSE}
source(file= "1_DataShaping.R",  echo= FALSE)
source(file= "2_Analysis.R",     echo= FALSE)
source(file= "3_Deliverables.R", echo= FALSE)
```

Operating systems are themselves programming environments.  Using the `system()` function, commands written in the corresponding operating system's language can be passed to the system for execution.  This includes executing full scripts written in that language.

```
## Rmarkdown does not currently capture the output from system calls, so this
  ## snippet does not display results.  However, I encourage you to try it out
  ## on your own.

## Using the Unix command "whoami" to determine user name under
	## currently logged in.
system("whoami")

## Using the Unix command "ls" to list the files in the current
	## working directory.  This is equivalent to list.files()
system("ls")
list.files() ## the equivalent R function      
```

### Exercise - Practicing Read/ Write

Generate a simulated dataset.  Use each read function to save the dataset to disk.  Use the file system manipulation functions to move the file to a new direction, rename it, and delete the file in the old location.  Use each write function to read the dataset back from disk.  Confirm that the dataset read back in is identical to the original dataset.

### Vocabulary Table for Lesson B3

In order to program effectively, you will need to memorize basic functions, operators, and constants.  Write each of the functions/operators/constants below on a flash card.  On the back of each card, write a succinct definition of what it does and a example of a line code you could enter into console that uses it.  Drill with these cards until you have memorized them.  Then drill again, coming up with a fresh example for each and testing that example in the console.

In order to understand what each function/operator/constant does, use the `help()` function to pull the documentation for it.  For example, `help("objects")` would pull up the documentation for the function `objects()`.  This document includes a description of what the function does ("Description" section), a list of all the arguments that can be given to the function ("Arguments" section), and examples of how to use the function ("Examples" section) at the bottom.  Only copy the definition or example from the documentation to your flash card if you absolutely understand what it does.  Otherwise, substitute your own.

The help documentation may be a difficult to read at first but keep practicing.  Over time, getting useful information from the documentation will become effortless.  Resist the impulse to do a Google search before you have consulted the documentation.  Google results can be of mixed quality - sometimes you will get a thoughtful, efficient solution, sometimes you will get a byzantine work-around that will teach you bad habits.

|Read/Write Functions |File Translation Packages  |File System Manipulation |Internet File Retrieval  |Interface Access
|:-		|:-		|:-		|:-
|`getwd()`		|`foreign`		|`dir.create()`		|`close()`				|`assign()`
|`load()`			|`openxlsx`		|`dir.exists()`		|`download.file()`|`get()`
|`read.csv()`	|`readstata13`|`file.copy()`		|`file.size()`		|`options()`
|`readLines()`|`rjson`			|`file.exists()`	|`object.size()`	|`source()`
|`save()`			|							|`file.remove()`	|`unzip()`				|`system()`
|`setwd()`		|							|`file.rename()`	|`url()`
|`write.csv()`|							|`list.files()`		|`zip()`
|`writeLines()`|						|